\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-01-26 document/XSGRB_sample.tex
Part Susanna and Paolo comments

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{Ricker2004}. With the two instruments, the
Burst Alert Telescope (BAT) \citep{Barthelmy2005} and the X-Ray Telescope (XRT)
\citep{Burrows2005}, \swift~is an ideal starting point for GRB investigations. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been greatly facilitated by the open data access policy of the
{\it Swift} mission.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or poorly localized bursts as reported by the alerting
telescope. To maximize the return of the follow-up campaign, we have chosen a
few selection criteria that facilitates an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of \textit{completeness} is desired \citep[e.g.,][]{Perley2016b}.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the \textit{Swift} bursts
	detected until the end of 2016. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red outline is added around GRBs which have been followed
	up by X-shooter in this work. Blue symbols with red outlines represent GRBs
	that enter our sample and have X-shooter spectroscopy. The different samples
	are compared in Sect. \ref{results}. The background shows the dust maps
	presented in \citet{Schlegel1998} where the sample criteria cut with $A_V
	\lesssim 0.5$ mag is replaced with a white color. The colorbar below indicates
	the value of $A_V$ on the plot. The dotted lines indicate intervals of
	30$^\circ$ in longitude and latitude.} \label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of the two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of 8 years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260 (PI:
Fynbo). We have included a few additional bursts, from the programmes
084.D-0265(A) (PI: Benetti), 091.A-0877(A) (PI: Schady), 092.D-0056(A) (PI:
Rau), 092.D-0633(E) (PI: Greiner), and 095.B-0811(B) (PI: Levan). The total
collection of spectra represents \textit{all} GRB afterglows that have been
followed up by X-shooter up to \termdate.

@@

Before any reductions are initiated, the raw object images are run through the
cosmic-ray removal algorithm \citep{VanDokkum2001} implementation,
\textit{Astro-SCRAPPY} \footnote{\url{https://github.com/astropy/astroscrappy}},
where a wide clipping radius have been used around detected cosmics to ensure that edge residuals are
robustly rejected.

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, initial sky-subtraction and lastly the rectification and merging
of the orders. For the initial sky-subtraction, the background has been
estimated in regions adjacent to the object trace clear of contaminating
sources. Because of the broken ADC, for some objects there is a lot of curvature
in the object trace along the dispersion axis of the slit. This means that for
some bursts, the initial sky-estimate has been made from a limited number of
pixels in the spatial direction. By doing an initial subtraction of the sky
background on the un-rectified image we ensure that the bulk of the sky
background is not redistributed by the rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.06 nm/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, we choose a sampling of 0.02 nm/pixel to ensure
that the bluest part of neither of the arms have a sampling lower than the
Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it is important to estimate the background variance in a large enough
region, so that any correlation between the signal and the weights are removed.
To this end, the weight map is generated by a running median window over the
variance map, where the trace has been masked and width of the window is chosen
to be wide enough for median variance to be generated on the basis of several
hundred pixels. This weighting scheme automatically also optimally combines
images of different exposure times or images where the background is varying,
which is often the case when a burst has been observed close to twilight

@@

By STARE reducing all observations we additionally get a spectrum of the sky,
which we can use to calibrate the wavelength solution.

@@

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
\citep{Smette2015, Kausch2015}, which can then be used to correct for the
absorption. The accuracy of the correction depends on the S/N per pixel for the
target spectrum, where we have chosen the requirement that the afterglow
continuum spectrum has a median signal-to-noise higher than a value of 10.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have overplot the most
	prominent lines seen in GRB afterglows from \citet{Christensen2011a}.
	Additionally, three intervening systems are seen in this sightline. This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode. In most cases there are therefore little
possibility to tweak slit widths to the seeing at the time of observations (i.e.
to optimize spectral resolution and signal to noise), and almost all our data is
therefore taken with a fixed set of slit widths and binning, described above. In
a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for host
galaxies). The delivered resolution for slit-width-dominated spectra
post-reduction and extraction can easily be determined from the bright sky
emission lines. For afterglow spectra with very high signal to noise, the
delivered spectral resolution can at times be determined from the science data
themselves. However, in the presence of multiple velocity components in
absorption, other forms of line broadening, and a lack of lines at some
redshifts, this is difficult to do at poorer signal to noise ratios (the
majority of spectra in our sample). A broad starting value for the expected
resolution will help fitting of these spectra, and can be important in upper
limit determination, and for this reason we construct a crude relation between
the seeing and the delivered resolution at our slit width, binning, and
reduction pipeline settings.

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to poor signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is perfect (the spatial profile of the trace is not a perfect
Gaussian), it nevertheless gives a sufficiently accurate estimate for the
analysis of these poor signal to noise science spectra.

@@

Out of the 158 bursts in the statistical sample, 33( $\sim$ 20 per cent) have
not been observed due to reasons unrelated to the GRB or afterglow properties.
The reasons include unavailability of the telescope due to technical maintenance
(e.g., mirror re-coating), a visitor rejecting the TOO trigger, or bad weather.
Because this cut is unrelated to the GRB properties, it will not change the
statistical properties of the full sample. Removing all these bursts,
drastically improves the redshift completeness from 61 per cent to 84 per cent.
The remaining burst not followed up already had redshift from other instruments
or were very faint and thus observations unlikely to yield anything. In the
remainder of the text, we consider the 125 bursts our statistical sample.

@@

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/betaOX.pdf}} \caption{Optical
	extinction relative to X-ray against X-ray derived hydrogen column densities.
	In red is shown the sample presented in \citet{Fynbo2009} where the lines
	indicate the kernel density estimate of the distribution. For the kernel
	density estimate, the limits have been replaced with values. We note that
	contrary to what is presented in \citet{Fynbo2009}, we exclude all bursts for
	which we do not have a redshift. Darker colors represented higher density of
	points. The corresponding marginal distributions are shown. In blue is the
	points for the bursts presented here along with the marginal histograms. Limits
	on $\beta_{OX}$ is shown by downwards facing arrows.} \label{fig:betaOX}
\end{figure}

It is clear, however, that the fraction of dark bursts for which we have
measured redshifts for is lower than the intrinsic dark burst fraction. To
measure a redshift, we either need a detected afterglow to obtain a spectrum or
to locate the host galaxy and determine the redshift from there. It is more
difficult to correctly associate a galaxy with a burst when there is no detected
optical afterglow and hence a correct redshift measurement is more difficult to
make, see \citet{Jakobsson2005, Levesque2010} and \citet{Perley2017} for a
recent example.

Using the table maintained by
Greiner\footnote{http://www.mpe.mpg.de/~jcg/grbgen.html}, we can see how the
presence of an optical afterglow affects the follow-up statistics. 50.5 per cent
of all \textit{Swift}-detected bursts do not have an detected optical afterglow,
but this number also includes bursts where no optical observations were
available, so is a lower limit. For the bursts entering our statistical sample,
the number of bursts without a detected optical afterglow is $\sim$ 27 per cent,
close to the upper limit on the fraction of dark bursts found in a complete
sample with a very high degree of redshift completeness \citep{Melandri2012}. Of
the bursts for which follow-up has actually been attempted, this number is
$\sim$ 18 per cent, confirming that we are indeed biased against bursts without
a detected optical afterglow in the spectroscopic sample. Since \textit{all}
GRBs should have an optical afterglow, the presence of an optical afterglow is
more a statement about the sensitivity than optical properties of the bursts.

Regardless that the fraction of dark bursts is lower in the observed sample,
compared to the statistical sample, the X-ray properties not differ
significantly, as shown in Sect. \ref{completeness}. This is despite
spectroscopic follow-up only being carried out in cases where either a
detectable optical afterglow or a clear counterpart are seen, which naively
should be biased against dark bursts occurring in more obscured galaxies, which
is shown to exhibit different galactic properties \citep{Perley2009,
	Kruhler2011, Rossi2012, Perley2013b, Perley2015b}. If the tendency for dark
bursts to have a higher $N_{\mathrm{HI, X}}$ is real, that the decreased
fraction of dark bursts does not alter the observed prompt X-ray brightness
distribution, potentially reflects the independence of the X-ray brightness on
the density of the circumburst medium \citep{Freedman2001, Berger2003,
	Nysewander2009}, if the measured $N_{\mathrm{HI, X}}$ is primarily driven by a
metal column local to the burst. Because we only use values for $N_{\mathrm{HI,
		X}}$ in the comparison for which the GRB has a measured redshift, this measure
is likely biased toward optically brighter bursts.

@@

Because the locations of long GRBs are associated with intensely star forming regions
\citep{Hogg1999, Bloom2002, Fruchter2006}, the GRB afterglow light has to
propagate through the large amounts of hydrogen fueling the star formation.
Because a significant fraction of the hydrogen has not been ionized yet, the
optical depth at the wavelength of \lya~is very high, saturating not only the
line center, but also the damping wings. This causes a strong absorption system
to appear in the afterglow continuum. For bursts with $z \gtrsim 1.7$, the
position of \lya~moves into the spectroscopic coverage of X-shooter, meaning
that we are able to detect this absorption trough due to \lya.

@@

In the compilation of $N_{\mathrm{HI}}$ measurements towards GRBs in Tanvir et
al. (submitted) %\citet{Tanvir2017}
there are 81 published $N_{\mathrm{HI}}$ values. We here provide 39 new neutral
hydrogen column density measurements -- an increase of the number of optically
derived hydrogen column densities of $\sim$ 50 per cent. We compare the median,
the 14th, and 86th percentiles of the two distributions. The sample presented
has a \nh~$= 21.8_{-0.8}^{+0.3}$ and the rest of the literature values has
\nh~$= 21.5_{-1.5}^{+0.4}$. We see that the two distribution has a large degree
of overlap due to the large width of the distributions, but we find a slightly
higher median value for the new sample presented here. A 2-sided KS test gets us
a p-value of p = 0.006, meaning relatively large evidence against the null that
the two samples are drawn from the same underlying distribution. Because the
bursts that have measurements of the hydrogen column density are selected solely
based on our ability to infer a column, it is difficult to make any strong
conclusions about the population statistics in terms of gas content.

We show the comparison between the two samples in Fig. \ref{fig:NH_dist}. We
additionally show the column density-distribution for the 12081 quasar absorbers
with \nh~$> 20$ from \citet{Noterdaeme2012b}. The fact that GRBs systematically
are located behind the highest \nh, previously noted by
\citep[e.g.,][]{Prochaska2007, Fynbo2009}, is stricking. The reason for this is
that quasar sample sight-lines through galaxies that are cross-section selected,
whereas GRB sight-lines probe the dense, star-forming regions in their hosts.

\begin{figure}
	\centering \includegraphics[width=9cm]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citep{Noterdaeme2012b} in blue. Overplot in
	green is the kernel density estimate is the similar distribution, only for
	absorbers in GRB sightlines. Values are taken from the compilation in Tanvir et
	al. (submitted), along with the new values presented in this sample. The
	marginal distributions for the two distribution are also shown along the left
	side and on the top, where the different environments probed are clearly
	visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}.} \label{fig:NH_dist}
\end{figure}

@@

\subsection{GRB~120211A (z = 2.346)}
The data presented here has been published in \citet{Kruhler2015}. Two
observations of the host of GRB~120211A has been obtained, starting 2013.02.17,
$> 1 year$ after the burst has faded. A redshift for this object has been
reported by \citet{Kruhler2015} and the features seen by those authors are
reproduced in these reductions, confirming $z =	2.346$.

>>>>

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{VanParadijs2000, Ricker2004}. With the two
on-board instruments, the Burst Alert Telescope (BAT) \citep{Barthelmy2005} and
the X-Ray Telescope (XRT) \citep{Burrows2005}, \swift~is an ideal telescope for
GRB hunting. A crucial aspect of the success of the {\it Swift} mission has been
the extensive ground based follow-up observations of the afterglows and of the
host galaxies of the GRBs, involving a large community of researchers. This
fruitful collaboration has been greatly facilitated by the open data access
policy of the {\it Swift} mission. The close collaboration between detection
facilities and electromagnetic  follow-up campaigns continue to be immensely
rewarding, as recently highlighted by the simultaneous detection of
gravitational waves and light from the neutron star merger in the shape of
GW170817/GRB180817A/AT2017gfo \citep{LIGOScientificCollaboration2017a,
	LIGOScientificCollaboration2017}.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or unconstrained burst localizations as reported by the alerting
telescope. To maximize the return of the follow-up campaign, we have chosen a
few selection criteria that facilitates an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of \textit{completeness} is desired \citep[e.g.,][]{Perley2016b}.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. Star symbols
	indicate bursts with measured redshifts and the circles indicate those without.
	The empty stars/circles are the positions of all the \textit{Swift} bursts
	detected until the end of 2016. A blue filling of the symbol indicates the
	position of bursts fulfilling the sample criteria specified in Sect.
	\ref{samplecrit}. The red outline is added around GRBs which have been followed
	up by X-shooter in this work. Blue symbols with red outlines represent GRBs
	that enter our sample and have X-shooter spectroscopy. The different samples
	are compared in Sect. \ref{results}. The background shows the dust maps
	presented in \citet{Schlegel1998}. Please note that we removed the background
	where the sample criteria cut ($A_V \lesssim 0.5$ mag) and replaced it with a
	white color visible at the equator. The colorbar below indicates the value of
	$A_V$ on the plot. The dotted lines indicate intervals of 30$^\circ$ in
	longitude and latitude.} \label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of the two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of 8 years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260 (PI:
Fynbo). These proposals were initiated on Guaranteed Time. We have included a
few additional bursts, from the programmes 084.D-0265(A) (PI: Benetti),
091.A-0877(A) (PI: Schady), 092.D-0056(A) (PI: Rau), 092.D-0633(E) (PI:
Greiner), and 095.B-0811(B) (PI: Levan). The total collection of spectra
represents \textit{all} GRB afterglows that have been followed up by X-shooter
up to \termdate.

@@

Before any reductions are initiated, the raw object images are run through the
cosmic-ray removal algorithm \citep{VanDokkum2001} implementation,
\textit{Astro-SCRAPPY}\footnote{\url{https://github.com/astropy/astroscrappy}},
where a wide clipping radius have been used around detected cosmics to ensure that edge residuals are
robustly rejected.

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, sky-subtraction and lastly the rectification and merging
of the orders. Errors and bad pixel maps are propagated throughout the
extraction. For the initial sky-subtraction, the background has been estimated
by a running median in regions adjacent to the object trace clear of
contaminating sources. Due to the broken ADC, for some objects there is
curvature in the object trace along the dispersion axis of the slit. This means
that for some bursts, the initial sky-estimate has been made from a limited
number of pixels in the spatial direction. By doing a subtraction of
the sky background on the un-rectified image we ensure that the bulk of the sky
background is not redistributed by the rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.06 nm/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, we choose a sampling of 0.02 nm/pixel to ensure
that the lowest wavelength part of neither of the arms have a sampling lower
than the Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it is important to estimate the background variance in a large
enough region, so that any correlation between the signal and the weights are
removed. To this end, the weight map is generated by a running median window
over the variance map produced by the pipeline, where the trace has been masked
and width of the window is chosen to be wide enough for median variance to be
generated on the basis of several hundred pixels. This weighting scheme
automatically also optimally combines images of different exposure times or
images where the background is varying, which is often the case when a burst has
been observed close to twilight

An additional sky-subtraction is additionally run on all rectified 2D spectra.
This is done to remove residual sky, still present. At each pixel in the
dispersion direction, the spatial dimension is fit with a low-order polynomial,
after all sources are masked out. This low order polynomial is then convolved
with a few pixel wide Gaussian filter in the dispersion direction and subtracted
of the entire 2D image.

@@

Using this method for all observations we additionally get a spectrum of the
sky, which we can use to calibrate the wavelength solution.

@@

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
(\texttt{Molecfit} \citep{Smette2015,
	Kausch2015}\footnote{\url{http://www.eso.org/sci/software/pipelines/skytools/molecfit}}), which can then be used to correct for the absorption. The accuracy of the correction depends on the S/N per pixel for the target spectrum, where we have chosen the requirement that the afterglow continuum spectrum has a median signal-to-noise higher than a value of 10.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$,
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$r$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have overplot the most
	prominent lines seen in GRB afterglows from \citet{Christensen2011a}. The
	regions of most severe telluric absorption is highlighted by a grey-shading the
	background. Additionally, three intervening systems are seen in this sightline.
	This spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode. In most cases there are therefore little
possibility to tweak slit widths to the seeing at the time of observations (i.e.
to optimize spectral resolution and signal to noise), and almost all our data is
therefore taken with a fixed set of slit widths and binning, described above. In
a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for host
galaxies). The delivered resolution for slit-width-dominated spectra
post-reduction and extraction can easily be determined from the bright sky
emission lines. For afterglow spectra with very high signal to noise, the
delivered spectral resolution can at times be determined from the science data
themselves. However, in the presence of multiple velocity components in
absorption, other forms of line broadening, and a lack of lines at some
redshifts, this is difficult to do at lower signal to noise ratios (the
majority of spectra in our sample). A broad starting value for the expected
resolution will help fitting of these spectra, and can be important in upper
limit determination, and for this reason we construct a crude relation between
the seeing and the delivered resolution at our slit width, binning, and
reduction pipeline settings.

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to low signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. While this
simple method is perfect (the spatial profile of the trace is not a perfect
Gaussian), it nevertheless gives a sufficiently accurate estimate for the
analysis of these low signal to noise science spectra.

@@

Out of the 158 bursts in the statistical sample, 33( $\sim$ 20 per cent) have
not been observed due to reasons unrelated to the GRB or afterglow properties.
The reasons include unavailability of the telescope due to technical maintenance
(e.g., mirror re-coating), a visitor rejecting the TOO trigger, or bad weather.
Because this cut is unrelated to the GRB properties, it will not change the
statistical properties of the full sample. Removing all these bursts for the
statistical sample, drastically improves the redshift completeness from 61 per
cent to 84 per cent. The remaining burst not followed up already had redshift
from other instruments or were very faint and without a host association thus
observations unlikely to yield a redshift measurement. In the remainder of the
text, we consider the 125 bursts our statistical sample.

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=9cm]{figures/betaOX.pdf}} \caption{Optical
	extinction relative to X-ray against X-ray derived hydrogen column densities.
	In red is shown the sample presented in \citet{Fynbo2009} where the lines
	indicate the kernel density estimate of the distribution. For the kernel
	density estimate, the limits have been replaced with values. We note that
	contrary to what is presented in \citet{Fynbo2009}, we exclude all bursts for
	which we do not have a redshift. Darker colors represented higher density of
	points. The corresponding marginal distributions are shown. In blue is the
	points for the bursts presented here along with the marginal histograms. Limits
	on $\beta_{OX}$ is shown by downwards facing arrows.} \label{fig:betaOX}
\end{figure}

Using the table maintained by
Greiner\footnote{http://www.mpe.mpg.de/~jcg/grbgen.html}, we can see how the
presence of an optical afterglow affects the follow-up statistics. 50.5 per cent
of all \textit{Swift}-detected bursts do not have a detected optical afterglow,
but this number also includes bursts where no optical observations were
available, so is a lower limit. For the bursts that enter our statistical
sample, the percentage of bursts without a detected optical afterglow is 28 per
cent, close to the upper limit on the fraction of dark bursts found in a
complete sample with a very high degree of redshift completeness
\citep{Melandri2012}. Of the bursts for which follow-up has actually been
attempted, this number is 23 per cent, suggesting a slight bias against
bursts without a detected optical afterglow in the spectroscopic sample.

However, the fraction of dark bursts for which we have measured redshifts for is
lower than the for the ones with a detected optical afterglow. For the
afterglows we have observed as part of the statistical sample that do not have a
detected optical afterglow, the redshift completeness is 53 per cent. For
comparison, for the afterglows in the statistical sample we have observed with
an optical afterglow detected, the redshift completeness is 91 per cent. This
also goes to show that the lack of redshift completeness in the sample presented
here is in part due to the increased difficulty of measuring a redshift for
bursts without an afterglow. To measure a redshift, we either need a detected
afterglow to obtain a spectrum or to locate the host galaxy and determine the
redshift from there. It is more difficult to correctly associate a galaxy with a
burst when there is no detected optical afterglow and hence a correct redshift
measurement is more difficult to make, see \citet{Jakobsson2005, Levesque2010}
and \citet{Perley2017} for a recent example.

Regardless that the fraction of dark bursts is lower in the observed sample,
compared to the statistical sample, the X-ray properties not differ
significantly, as shown in Sect. \ref{completeness}. This is despite
spectroscopic follow-up only being carried out in cases where either a
detectable optical afterglow or a clear counterpart are seen, which naively
should be biased against dark bursts occurring in more obscured galaxies, which
is shown to exhibit different galactic properties \citep{Perley2009,
	Kruhler2011, Rossi2012, Perley2013b, Perley2015b}. That the decreased fraction
of dark bursts in observed sample does not alter the observed prompt X-ray
brightness distribution, potentially reflects the independence of the X-ray
brightness on the density of the circumburst medium \citep{Freedman2001,
	Berger2003, Nysewander2009}, if the measured $N_{\mathrm{HI, X}}$ is primarily
driven by a metal column local to the burst. Because we only use values for
$N_{\mathrm{HI, X}}$ in the comparison for which the GRB has a measured
redshift, this measure is likely biased toward optically brighter bursts.

@@

The locations of long GRBs are associated with intensely star forming regions
\citep{Hogg1999, Bloom2002, Fruchter2006}. Because a significant fraction of the
hydrogen in these regions has not been ionized yet, the optical depth at the
wavelength of \lya~is very high, saturating not only the line center, but also
the damping wings. This causes a strong absorption system from the
\lya-transition to appear in the afterglow continuum. For bursts with $z \gtrsim
1.7$, the position of \lya~moves into the spectroscopic coverage of X-shooter,
meaning that we are able to detect this absorption trough due to \lya.

@@

In the compilation of $N_{\mathrm{HI}}$ measurements towards GRBs in Tanvir et
al. (submitted) %\citet{Tanvir2017}
there are 81 published $N_{\mathrm{HI}}$ values, excluding the measurements
provided here. We here provide 39 new neutral hydrogen column density
measurements -- an increase of the number of optically derived hydrogen column
densities of $\sim$ 50 per cent. We show the two distributions in  Fig.
\ref{fig:NH_dist}. We compare the median, the 14th, and 86th percentiles of the
two distributions. The sample presented has a \nh~$= 21.8_{-0.8}^{+0.3}$ and the
rest of the literature values has \nh~$= 21.5_{-1.5}^{+0.4}$. We see that the
two distribution has a large degree of overlap due to the large width of the
distributions, but we find a slightly higher median value for the new sample
presented here. A 2-sided KS test gets us a p-value of p = 0.006, meaning
relatively large evidence against the null that the two samples are drawn from
the same underlying distribution. Because the bursts that have measurements of
the hydrogen column density are selected solely based on our ability to infer a
column, it is difficult to make any strong conclusions about the population
statistics in terms of gas content.

In Fig. \ref{fig:NH_dist}, we additionally show the column density-distribution
for the 12081 quasar absorbers with \nh~$> 20$ from \citet{Noterdaeme2012b}. The
fact that GRBs systematically are located behind the highest \nh, previously
noted by \citep[e.g.,][]{Prochaska2007, Fynbo2009}, is very clear in this
figure. The reason for this is that quasar sample sight-lines through galaxies
that are cross-section selected, whereas GRB sight-lines probe the dense,
star-forming regions in their hosts.

\begin{figure}[!ht]
	\centering \includegraphics[width=9cm]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citep{Noterdaeme2012b} in blue. Overplot in
	green is the kernel density estimate is the similar distribution, only for
	absorbers in GRB sightlines. Values are taken from the compilation in Tanvir et
	al. (submitted), along with the new values presented in this sample. The
	marginal distributions for the two distribution are also shown along the left
	side and on the top, where the different environments probed are clearly
	visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}.} \label{fig:NH_dist}
\end{figure}

@@

\subsection{GRB~120211A (z = 2.346)}
The data presented here has been published in \citet{Kruhler2015}. Two
observations of the host of GRB~120211A has been obtained, starting 2013.02.17,
$> 1$ year after the burst has faded. A redshift for this object has been
reported by \citet{Kruhler2015} and the features seen by those authors are
reproduced in these reductions, confirming $z =	2.346$.
